---
title: "NYC Flights Delays"
author: "Turki Alsaedi, Saad AlOtaibi, Yousef Alotaibi, Maan Al Neami, Shaimaa Alghamdi<br><br> Table of Contents"
output:
  html_document:
    df_print: paged
    toc: true
editor_options:
  chunk_output_type: inline
  
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../"
    )
  })
---
```{=html}
<style>
#caption{
    color: #777777;
}
body {
text-align: justify;
max-width: 800px;
margin: auto;
font-size: 18px;
}

table{
  position:relative;
  width: 160% !important;
  left: 50%;
  transform: translateX(-50%);
}

</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,    # hide code blocks
                      warning = FALSE, # hide warning messages
                      message = FALSE) 
library(here)
library(tidyverse)
library(knitr)
library(nycflights13)
library(plotly)
library(ggplot2)
library(gridExtra)
```

### Introduction
<br>

<img src = "../../delay-1.jpg">

<br>

Airport delays are one of the most common problems people face upon travelling. These delays are usually associated with certain carriers or destinations. Thus the question is, are they really the cause of these delays or is it due to other reasons?

In 2013, data about 336,776 flights were collected to answer this question, the data collected were about flights departing from New York City across all of its airports: John F. Kennedy International Airport (JFK), Newark Liberty International Airport (EWR) and LaGuardia Airport (LGA) to destinations all over the United States and some of its territories (Puerto Rico, and the American Virgin Islands). The data was described by 19 attributes shown in Table 1 followed by a sample of the data in Table 2.

<br>

#### Goal
The aim behind this analysis is to confirm whether there is a relationship between New York City’s flight delays and attributes of the dataset. We hypothesize that flight distance and destination are major contributors to the delays and that different destinations with different distances will have less delays.

<br>

#### Methods
- Data Visualization
- Exploratory Data Analysis (EDA)
- Data Munging

<br>

#### The Dataset

```{r}
flights_data <- read.csv(here("data/flights_data.csv"))

kable(flights_data)
```
<p id="caption">Table 1: A list of all attributes and its descriptions that are present in the dataset</p>

<br>

```{r}
library(reactable)
reactable(sample_n(flights, 5))
```
<p id="caption">Table 2: A sample of the dataset and how it is structured</p>
<br><br>

Now lets take a look at our data:

```{r}
flights_df <- flights

# Checking for null values.
flights <- flights_df %>% drop_na()
colSums(is.na(flights_df))

glimpse(flights_df)

summary(flights_df)

```
<br><br>

### EDA

Although 3 airports per city may seem large, larger than many of the world’s capitals, the city that never sleeps hosts only 3 out of 16 airports of the state of New York which itself is not the host of the largest number of airports per state as shown in Figure 1.

<br>

```{r}
airports_per_state <- read_csv(here("data/airports_per_state.csv"))

airports_per_state %>% 
  group_by(state, state_abv) %>% 
  summarise(airports_no = n()) -> airports_per_state

# Plotting the map

fig <- plot_geo(airports_per_state, locationmode = 'USA-states')

# specifying scale and options
g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)

fig %>%
  add_trace(
    z = ~airports_no, text = ~state, locations = ~state_abv,
    color = ~airports_no, colors = 'Blues') %>% 
  colorbar(title = "Airports") %>%
  layout( title = 'No. of airports per State',
          geo = g)
```

<p id="caption">Figure 1: An interactive map showing the number of airports per state. Puerto Rico (7 airports) and U.S. Virgin Islands (2 airports) are not shown.</p>

Based on the figure we can observe that:

- Alaska has 28 airports making it the state with the largest number of airports followed by Texas (24) and California (23).
- Only one state (Vermont) has one airport.
- Every state has at least one airport.

<br>
The three airports of the city of New York serve different purposes–such as domestic or international–and hence should have different number of flights, let us confirm that using Figure 2.

```{r}
flights %>% 
  group_by(origin) %>% 
  summarise(count_flights = n()) %>% # % of flights
  ggplot(aes(x= origin, y = count_flights)) +
  geom_col(fill="light blue") +
  labs(x = "NYC Airport", y = "No. of flights")
```
<p id="caption">Figure 2: Number of flights per NYC airport</p>

From the figure above, we can see that all of the airports have relatively similar numbers with LaGuardia Airport (LGA) having the lowest, probably due the fact that it is a domestic airport. Newark Liberty International Airport (EWR) has the largest share of flights, which can be explained by looking at its location, where it lays on the border between New York state and New Jersey state, making it a strategic location and more favourable over JFK International Airport.

<br>
And despite the large number of flights and airports around the United States, Figure 3 showed that some states were never reached from NYC airports during 2013.

<br>

```{r}
airports_per_state <- read_csv(here("data/airports_per_state.csv"))

airports_per_state %>% 
  rename(dest = airport_code)-> airports_per_state

left_join(flights, airports_per_state, by='dest') -> flights

flights %>%
  group_by(state, state_abv) %>% 
  summarise(flights_per_state = n()) -> flights_map_df

fig <- plot_geo(flights_map_df, locationmode = 'USA-states')

g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)

fig %>%
  add_trace(
    z = ~flights_per_state, text = ~state, locations = ~state_abv,
    color = ~flights_per_state, colors = 'YlGnBu') %>% 
  colorbar(title = "Flights per state") %>%
  layout( title = 'No. of flights to destination by State',
          geo = g)

```
<p id="caption">Figure 3: Map showing flights per state as destination</p>

The map shows us that the most visited destination is Florida, followed by California by almost half of the number. The map also shows that there are 8 states with zero flights towards it namely: Mississippi, Kansas, Idaho, New Hampshire, New Jersey, Delaware, South and North Dakota.

<br>
Figure 4 will show us the delays per airport:
<br>
```{r}
flights %>% 
  select(origin, arr_delay, dep_delay) %>% 
  pivot_longer(-c(origin)) -> plot4

posn_d <- position_dodge(width = 0.3)

# ggplot(head(plot4, 200), aes(x=origin, y=value, color=name)) + # for shoter interpretation time
ggplot(plot4, aes(x=origin, y=value, color=name)) +
  geom_point(position = posn_d) +
  stat_summary(fun = mean, na.rm = TRUE, 
               geom = "point", color = "black", size = 2) +
  labs(x="Airport", y="Delay (minutes)", color = "Delay type") +
  scale_color_brewer(palette = "Accent",
                     labels = c("Arrival delay", "Departure delay")) +
  stat_summary(fun=mean, colour="black", geom="text", show.legend = FALSE, 
                 vjust=-0.7, aes( label=round(..y.., digits=1)))
```
<p id="caption">Figure 4: Number of delays per NYC airport</p>
The figure features positive and negative points–indicating departure/arrival was before time or ahead of scheduled time. The delay average across all airports is +9.7 minutes. Although each airport shows a different average, it is not enough to say that one airport will have a certain delay time.

To further investigate that, let us take a look at the attributes and how they affect each other.


```{r}
all_df <- subset(flights,
                 select = c(dep_time, sched_dep_time, dep_delay,
                            arr_time, sched_arr_time, arr_delay,
                            air_time, distance) )

all_matrix <- cor(all_df, use = "complete.obs")

library(corrplot)
corrplot(all_matrix,
         type = "upper",
         order = "hclust",
         tl.col = "black",
         tl.srt = 45,
         diag = FALSE,
         method = "number",
         col = COL2('PRGn'))
```
<p id="caption">Figure 5: Correlation matrix of the dataset attributes</p>

Here we see the correlation matrix between the main numeric attributes.

- Surprisingly there is no correlation at all between `distance` and the delays (`dep_delay` and `arr_delay`) which was the opposite of what we expected.
- There is no correlation between `air_time` (time airplane spent in air) and the delays.
- There is a *strong* correlation between departure delay and arrival delay which is reasonable since one is caused by the other.
- There is some obvious *strong* correlations like `distance` and `air_time` and the bottom between the attributes: departure time, scheduled departure time, arrival time, scheduled arrival time which is expected. 

<br>

Although the correlation matrix showed us the relationship between many variables, it did not mention a very important aspect of flights, time.
<br><br>

```{r}
flights_df %>%
  filter(dep_delay > 0) %>%
  count(hour, dep_delay) %>% 
  ggplot(aes(hour, dep_delay)) +
  geom_col() +
  labs(x = "Hour of day",
       y = "Number of Departure delays",
       title = "Number of Departure delay by hour of days") -> dep_plot1

flights_df %>%
  filter(arr_delay > 0) %>%
  count(hour, arr_delay) %>% 
  ggplot(aes(hour, n)) +
  geom_col() +
  labs(x = "Hour of day",
       y = "Number of Arrival delays",
       title = "Number of Arrival delays by hour") -> arr_plot1

grid.arrange(dep_plot1, arr_plot1, ncol = 2)
```
<p id="caption">Figure 6: Departure and ARRIVAL delays per hour</p>

<br>

The first plot on the left indicate that the lowest number of departure delays occur on the early hours of day at 5 am, and late at night at 10-11pm. And the worst time is between 4-7 pm

<br>

The second plot on the right indicate similar result to the previous one, with the lowest number of arrival delays occur on the early hours of day at 5 am, and late at night at 10-11pm. And the worst time is between 4-7 pm

<br>


``` {r}
flights %>%
  filter(dep_delay > 0) %>%
  count(month, dep_delay) %>% 
  ggplot(aes(month, n)) +
  geom_col() +
  labs(x = "Month",
       y = "Number of Departure delays",
       title = "Number of Departure delays by month") +
  xlim(0,12.5) -> dep_plot2

flights %>%
  filter(arr_delay > 0) %>%
  count(month, arr_delay) %>% 
  ggplot(aes(month, n)) +
  geom_col() +
  labs(x = "Month",
    y = "Number of Arrival delays",
    title = "Number of Arrival delays by month") +
  xlim(0,12.5) -> arr_plot2

grid.arrange(dep_plot2, arr_plot2, ncol = 2)

```
<p id="caption">Figure 7: Departure and Arrival delays per month</p>

<br>

We can see from the first plot that the highest number of departure delays occur on month 6 (Jun), 7 (Jul), and 12 (Dec), which indicate there are more departure delays during summer and winter breaks.

<br>
Similar to the other plot, in the second plot we see highest number of arrival delays occur on month 7 (Jul), and 12 (Dec) during the summer and winter breaks.

<br>

<br>
```{r}
flights %>%
  count(month, arr_delay, flight) %>%
  ggplot(aes(as.factor(month), fill = (arr_delay > 0))) +
  geom_bar(position = "stack") +
  labs(x = "Month",
       y = "Number of Flights",
       title = "Number of flights by month") +
  guides(fill = guide_legend("Arrival Delay"))
```
<p id="caption">Figure 8: Number of flights and their arrival delays month</p>

<br>
In the first plot on the left, we can see that as the number of flights increases, the number of departure delays also increase.
If we check for the months with the highest delays, 6 (Jun), 7 (Jul), and 12 (Dec). We see they also have the highest 
number of flights compared to other months. 

<br>
The second plot is similar to the first where the number of flights increases, the number of Arrival delays also increase. Except that for month 8 (Aug), where the number of flights were high but the arrival delays were relatively lower than months with less flights.

<br>

```{r, figures-side, fig.show="hold", out.width="50%"} 
# library("gridExtra")  

flights_df1 <- flights

# Categorizing the dep_delay & arr_delay time into 5 categories 

flights_df1$dep_delay_cat <- cut(flights_df1$dep_delay,
                     breaks = c(-45,-1,0,15,60,1301),
                     labels = c("Before Time",
                                "On Time",
                                "Low Delay (less than 15 min)",
                                "Medium Delay (16-60 min)",
                                "High Delay (more than 60 min)"))

flights_df1$arr_delay_cat <- cut(flights_df1$arr_delay,
                     breaks = c(-88,-1,0,15,60,1272),
                     labels = c("Before Time",
                                "On Time",
                                "Low Delay (less than 15 min)",
                                "Medium Delay (16-60 min)",
                                "High Delay (more than 61 min)"))

# Plotting dep_delay_cat and arr_delay_cat based on it's count in the year
# 2013, so we can know how much each represent in our sample.

ggplot(flights_df1, aes(dep_delay_cat)) +
  geom_bar(fill = "cadetblue3") +
  geom_text(aes(label= scales::percent(after_stat(as.double(prop))), group=1),
            stat='count', vjust = -0.3,) +
  scale_x_discrete(guide = guide_axis(n.dodge=2))+
  labs(title="Percentage of NYC Flights Departure Delay (2013)",
       x = "Departue Delay Time",
       y = "Number of Departure Delays")


ggplot(flights_df1, aes(arr_delay_cat)) +
  geom_bar(fill = "cadetblue3") +
  geom_text(aes(label= scales::percent(after_stat(as.double(prop))), group=1),
            stat='count', vjust = -0.3, ) +
  scale_x_discrete(guide = guide_axis(n.dodge=2)) +
  labs(title="Percentage of NYC Flights Arrival Delay (2013)",
       x = "Arrival Delay Time",
       y = "Number of Arrival Delays")

# grid.arrange(dep_plot, arr_plot, ncol = 2)

``` 
<p id="caption">Figure 9: Percentage of flights' departure and arrival delays</p>

We can see that almost 39% of the NYC flights in the year 2013 had a departure delay, only 5% departed on time and 55.9% departed before time. As for the arrival delays, we can see that it doesn't differ that much from the departure delays.

<br>

```{r} 

# Summarizing each carrier with no_flights and percentage of dep_delays.

flights_df1 %>% 
  group_by(carrier) %>%
  summarise(no_flights = n(), 
            low_delay = scales::percent(sum(dep_delay > 0 & dep_delay < 16)/no_flights),
            medium_delay = scales::percent(sum(dep_delay > 16 & dep_delay < 61)/no_flights),
            high_delay = scales::percent(sum(dep_delay > 60)/no_flights),
            overall_delay = scales::percent(sum(dep_delay > 0)/no_flights)) %>%
  arrange(-no_flights) %>% 
  kable()
```
<p id="caption">Table 3: Carriers, their number of flights and the percentage of flights with delays</p>

WN carriers tend to have the highest percentage of overall delays, and US carriers tend to have a low percentage of delays compared to its number of flights.

<br><br>
```{r}

# Plotting each carrier based on dep_delay_cat

flights_pos_delay <- flights_df1 %>%
  filter(dep_delay > -1)

ggplot(flights_pos_delay, aes(y = carrier, fill = dep_delay_cat)) +
  geom_bar() +
  labs(x=" Number of flights", y="Carrier", fill = "Delay type") +
  scale_color_brewer(palette = "Accent")
```
<p id="caption">Figure 10: Carriers, their number of flights and the percentage of flights with delays</p>

From the horizontal stacked bar it's clear that carriers UA, EV, B6 and DL have the highest frequency of delays, also these carriers have the highest number of flights, which can tell us that carriers having a high number of flights tend to have a high frequency of delays.

<br>
```{r}

# Summarizing each carrier with highest_delay and avg_delay

flights_df1 %>%
  group_by(carrier) %>%
  summarise(highest_delay = max(dep_delay), avg_delay = mean(dep_delay)) %>%
  arrange(-avg_delay) %>% 
  kable()
 
```
<p id="caption">Figure 11: Carriers and their maximum and average delays</p>

The carrier with the highest delay time is HA with 1301 min delay and the carrier with the highest avg delay is F9 with an average of 20.2 

<br>

### Conclusion
After conducting the EDA, we have found out that delays on average do not exceed 15 minutes. The source of the delays vary and is related to multiple factors, notably: date and time of the flight and the flight carrier.

<br>

### Resources
Wickham H (2022). nycflights13: Flights that Departed NYC in 2013. R package version 1.0.2, [https://github.com/hadley/nycflights13](https://github.com/hadley/nycflights13).

<br>

### Source Code
This report is hosted on Github Pages and the repo can be accessed via [this link](https://github.com/mhaln3mi/nyc-flights-EDA).

<br><br><br>
