---
title: "NYC Flights EDA"
date: '2022-06-22'
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../"
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Libraries
library(nycflights13)
library(tidyverse)
library(janitor)

# reading the dataset
flights -> flights_df

```

## Atrributes description 
+--------------------+---------------+----------------------------------------------+
|Variable            |Type           |    Description                               |
+====================+===============+==============================================+
| fixed_acidity      |    Double     | most acids involved with wine or fixed or    |
|                    |               | nonvolatile (do not evaporate readily)       |
+--------------------+---------------+----------------------------------------------+
|Voltaile_acidity    |    Double     | the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
| citric_acid        |    Double     | found in small quantities, citric acid can add 'freshness' and flavor to wines|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|residual_sugar      |    Double     | the amount of sugar remaining after fermentation stops|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|chlorides           |    Double     | the amount of salt in the wine               |
+--------------------+---------------+----------------------------------------------+
|free_sulfur_dioxide |    Double     | the free form of SO2 exists in equilibrium between molecular SO2| 
|                    |               | (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|total_sulfur_dioxide|    Double     | amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm,| 
|                    |               | SO2 becomes evident in the nose and taste of wine|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|density             |    Double     | the density of water is close to that of water depending on the percent alcohol and sugar content|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|pH                  |    Double     | describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|sulphates           |    Double     | a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|alcohol             |    Double     | the percent alcohol content of the wine      |
+--------------------+---------------+----------------------------------------------+
|quality             |    Double     | Score between (3-8)                          | 
+--------------------+---------------+----------------------------------------------+


```{r, echo=FALSE, include=FALSE}

glimpse(flights_df)
summary(flights_df)

# Checking for duplicates.
get_dupes(flights_df)

# Checking for null values.
flights_df <- flights_df %>% drop_na()
colSums(is.na(flights_df))

```



```{r} 
# yousef

 
```



```{r}
# saad 

```


``` {r}
# turki
```



``` {r}
# maan
flights_df %>%
  filter(dep_delay > 0) %>%
  count(month, dep_delay, flight) %>% 
  ggplot(aes(month, n)) +
  geom_col() +
  labs(x = "Month",
       y = "Number of Departure delays",
       title = "Number of Departure delays by month") +
  xlim(0,12.5)
```

<br>

We can see from the plot above hat the highest number of departure delays occur on month 6 (Jun), 7 (Jul), and 12 (Dec)
which indicate there are more departure delays during summer and winter breaks.

<br>

```{r}
flights_df %>%
  filter(arr_delay > 0) %>%
  count(month, arr_delay) %>% 
  ggplot(aes(month, n)) +
  geom_col() +
  labs(x = "Month",
    y = "Number of Arrival delays",
    title = "Number of Arrival delays by month") +
  xlim(0,12.5)
```

<br>

similar to the previous plot, here we see highest number of arrival delays occur on month 7 (Jul), and 12 (Dec)
during the summer and winter breaks.

<br>

```{r}
flights_df %>%
  count(month, dep_delay, flight) %>% 
  ggplot(aes(as.factor(month), fill = (dep_delay > 0))) +
  geom_bar(position = "stack") +
  labs(x = "Month",
       y = "Number of Flights",
       title = "Number of flights by month") +
  guides(fill = guide_legend("Departure Delay"))
```

<br>

In this plot we can see that as the number of flights increases, the number of departure delays also increase.
If we check for the months with the highest delays, 6 (Jun), 7 (Jul), and 12 (Dec). We see they also have the highest 
number of flights compared to other months. 

<b>

```{r}
flights_df %>%
  count(month, arr_delay, flight) %>% 
  ggplot(aes(as.factor(month), fill = (arr_delay > 0))) +
  geom_bar(position = "stack") +
  labs(x = "Month",
       y = "Number of Flights",
       title = "Number of flights by month") +
  guides(fill = guide_legend("Arrival Delay"))
```

<br>

Here, it's similar to the plot above where the number of flights increases, the number of Arrival delays also increase. Except that for month 8 (Aug), where the number of flights were high but the arrival delays were relatively lower than 
months with less flights.

## Conclusion 


                  
## Citation

#### [name](url)

