---
title: "NYC Flights Delays"
author: "Turki Alsaedi, Saad AlOtaibi, Yousef Alotaibi, Maan Al Neami, Shaimaa Alghamdi<br><br> Table of Contents"
output:
  html_document:
    df_print: paged
    toc: true
editor_options:
  chunk_output_type: inline
---
```{=html}
<style>
#caption{
    color: #777777;
}
body {
text-align: justify;
max-width: 800px;
margin: auto;
font-size: 16px;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(here)        # relative paths in rmd
library(tidyverse)
library(knitr)
library(nycflights13)
library(ggplot2)
```

### Introduction
Airport delays are one of the most common problems people face upon travelling. These delays are usually associated with certain carriers or destinations. Thus the question is, are they really the cause of these delays or is it due to other reasons?

In 2013, data about 336,776 flights were collected to answer this question, the data collected were about flights departing from New York City across all of its airports: John F. Kennedy International Airport (JFK), Newark Liberty International Airport (EWR) and LaGuardia Airport (LGA) to destinations all over the United States and some of its territories (Puerto Rico, and the American Virgin Islands). The data was described by 19 attributes shown in Table 1.

#### Goal

#### Methods

### EDA
#### The Dataset

```{r}
flights_data <- read.csv(here("data/flights_data_v2.csv"))

kable(flights_data)
```
<p id="caption">Table 1: A list of all attributes and its descriptions that are present in the dataset</p>

<br>
Although 3 airports per city may seem large, larger than many of the world’s capitals, the city that never sleeps hosts only 3 out of 16 airports of the state of New York which itself is not the host of the largest number of airports per state as shown in Figure 1.

```{r}

```


When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

### Conclusion
### Resources
### Source Code
