---
title: "NYC Flights EDA"
date: '2022-06-22'
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../"
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Libraries
library(nycflights13)
library(tidyverse)
library(janitor)

# reading the dataset
flights -> flights_df

```

## Atrributes description 
+--------------------+---------------+----------------------------------------------+
|Variable            |Type           |    Description                               |
+====================+===============+==============================================+
| fixed_acidity      |    Double     | most acids involved with wine or fixed or    |
|                    |               | nonvolatile (do not evaporate readily)       |
+--------------------+---------------+----------------------------------------------+
|Voltaile_acidity    |    Double     | the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
| citric_acid        |    Double     | found in small quantities, citric acid can add 'freshness' and flavor to wines|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|residual_sugar      |    Double     | the amount of sugar remaining after fermentation stops|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|chlorides           |    Double     | the amount of salt in the wine               |
+--------------------+---------------+----------------------------------------------+
|free_sulfur_dioxide |    Double     | the free form of SO2 exists in equilibrium between molecular SO2| 
|                    |               | (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|total_sulfur_dioxide|    Double     | amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm,| 
|                    |               | SO2 becomes evident in the nose and taste of wine|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|density             |    Double     | the density of water is close to that of water depending on the percent alcohol and sugar content|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|pH                  |    Double     | describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|sulphates           |    Double     | a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant|
|                    |               |                                              |
+--------------------+---------------+----------------------------------------------+
|alcohol             |    Double     | the percent alcohol content of the wine      |
+--------------------+---------------+----------------------------------------------+
|quality             |    Double     | Score between (3-8)                          | 
+--------------------+---------------+----------------------------------------------+


```{r, echo=FALSE, include=FALSE}

glimpse(flights_df)
summary(flights_df)

# Checking for duplicates.
get_dupes(flights_df)

# Checking for null values.
is.null(flights_df)

```



```{r} 
# yousef

 
```



```{r}
# saad 

```


``` {r}
# turki
```



``` {r}
# maan
```


## Conclusion 


                  
## Citation

#### [name](url)

